Chaotic mixing of tidal debris

### 2015-07-04 ms.tex
andreas and monica comments

% Affiliations
\newcommand{\colum}{1}
\newcommand{\adrn}{2}
\newcommand{\mich}{3}
\newcommand{\nyu}{4}
\newcommand{\cds}{5}
\newcommand{\mpia}{6}
\altaffiltext{\colum}{Department of Astronomy, 
		              Columbia University, 
		              550 W 120th St., 
		              New York, NY 10027, USA}
\altaffiltext{\adrn}{To whom correspondence should be addressed: adrn@astro.columbia.edu}
\altaffiltext{\mich}{Department of Astronomy, 
			   University of Michigan,
			   Ann Arbor, MI 48109, USA}
\altaffiltext{\nyu}{Center for Cosmology and Particle Physics,
                      Department of Physics, New York University,
                      4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\cds}{Center for Data Science,
                      New York University,
                      4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                     K\"onigstuhl 17, D-69117 Heidelberg, Germany}

@@

The morphological (density) evolution of the tidal debris therefore depends only on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances --- from $\approx10$ to $100~{\rm kpc}$ --- and progenitor masses --- from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass --- \citep[][]{ibata94, belokurov06,grillmair06a,grillmair06b}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to N-body simulations \citep[see, e.g., the introduction of][]{kuepper15}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. 
% internal question: This is all just about the methods -- do I need sentences or a paragraph on results?

@@

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the axisymmetric potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further in this work.

In this work, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles due to the chaotic diffusion of the individual orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time --- for a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic halos and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams. We propose a new measure for determining whether this small-scale frequency diffusion will be important for the morphological evolution of tidal debris with a given spread in frequencies or orbital properties over finite and dynamically relevant times.

@@

A \emph{regular orbit} is a set of such time series that can be transformed to a special set of canonical coordinates known as angle-action coordinates. In these coordinates, the position variables are angles, $\boldsymbol{\theta}$, that increase linearly with time with rates set by $N$ constant, fundamental frequencies, $\boldsymbol{\Omega} = (\Omega_1, ..., \Omega_N)$. The frequency of a Fourier component in the sum for any individual component of motion (the $\omega_k$ in Equation~\ref{eq:fourier}) are just linear, integer combinations of the fundamental frequencies, $\boldsymbol{\Omega}$ --- that is, for a regular orbit, any Fourier component frequency may be written
\begin{align}
	\omega_k &= \boldsymbol{n_k} \cdot \boldsymbol{\Omega} \label{eq:fourierfreq}
\end{align} %; \, (\boldsymbol{n}_k \in \mathbb{Z}^3)
where $\boldsymbol{n}_k$ is a vector of $N$ integers. The conjugate momentum coordinates --- the actions,  $\boldsymbol{J}$ --- are constants of motion. Even stronger, the actions are isolating integrals and any pair are in involution such that
\begin{equation}
	[J_i, J_j] = 0
\end{equation}
where $[\cdot,\cdot]$ is the Poisson bracket. This implies that for an $N$ dof system, a regular orbit has $N$ independent constants of motion and the motion is therefore restricted to an $N$-dimensional manifold embedded in the 2$N$ dimensional phase space. The topology of angle-action space is toroidal and any regular orbit in an $N$ dof Hamiltonian can be understood as motion on the surface of an $N$-torus. Each set of actions, $(J_1,...,J_N)$ (or frequencies), uniquely labels a torus, and regular orbits are sometimes referred to in terms of their orbital tori.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position \citep[e.g.,][]{goldstein80}. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

The frequencies of a generic orbit are typically incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is referred to as a resonant orbit. Resonant orbits are confined to a surface with lower dimensionality than the surface of an orbital torus, depending on the number of resonance relations obeyed: we refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. Uni-resonant orbits in a triaxial potential are confined to a 2D surface, and bi-resonant orbits are closed 1D curves. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with example resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all orbits are regular.

@@

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined, the orbit may behave nearly regular for long periods of time.

For small values of $\epsilon$, many regular orbits survive and only small chaotic regions are introduced, especially where resonances overlap \citep[see][]{chirikov60}. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Figure~\ref{fig:cartoons} (right panel) qualitatively shows this phenomenon --- the resonant and resonant-layer orbits may still be regular, but many or all of the conditionally-periodic orbits can be chaotic. As the perturbation strength increases, next the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \cite[for a more quantitative illustration of this transition from integrability to global chaos, see Figure~9 in][]{valluri98}.

@@

Chaotic orbits have no orbital actions because they are only strictly bound to their energy hypersurface (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, but rather the frequencies that describe the character of motion evolve with time. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable (within a factor of a few) to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can not change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are generically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. With the inclusion of baryonic physics and sub-grid prescriptions for energy input due to supernovae and other feedback mechanisms, the inner potential ($\lesssim$$0.1R_{\rm vir}$ for a $\approx$$10^{12}~\msun$ halo mass) typically becomes more spherical, though the magnitude of this reshaping depends on the particular merger history and star formation efficiency within a given halo \citep[e.g.,][though in Milky Way-like galaxies, baryonic disks will add non-sphericity to the total potential]{dubinski94,kazantzidis04, bryan13, butsky15}. It is less clear what happens to the outer halo.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{ c  c }
	         Parameter & Value\\\toprule
		$v_c$ & 175~km~s$^{-1}$\\
		$r_s$ & 20~kpc\\
		$a$ & 1\\
		$b$ & 0.77\\
		$c$ & 0.55\\
		\bottomrule
		\end{tabular}
	\caption{Summary of parameters for the triaxial NFW potential (Equation~\ref{eq:potential}) used in this work. Triaxiality is introduced in the density (rather than the potential) to ensure that the density is physical at all radii. Velocity scale, scale radius, and axis ratios are chosen to match the median halo parameters for a $M_{\rm vir} \approx 10^{12}~\msun$ halo from \citep{jing02}. \label{tbl:potential}}
\end{center}
\end{table*}

@@

In Section~\ref{sec:results1}, we generate grids of orbits in the potential described above to map the orbit structure of the potential. We classify each orbit in terms of the strength of chaos along the orbit as computed using the Lyapnov and frequency diffusion times of Section~\ref{sec:methods}. With this initial classification, in Section~\ref{sec:results2} we follow the evolution of ensembles of orbits generated around each orbit in the initial grid. We find that the configuration-space density of ensembles around weakly chaotic orbits evolve faster (e.g., in mean density) than expected given the timescales over which chaos is computed to be relevant for the parent orbits. In Section~\ref{sec:results3}, we explain this phenomenon in the context of how chaotic diffusion occurs (e.g., Section~\ref{sec:behavior-chaotic}).

@@

Figure~\ref{fig:icmap} shows the grid of initial conditions in the $xz$ plane --- each pixel in this grid represents an orbit, and the major orbit families are labeled. Points are colored (black / white) whether the orbit acts like a tube orbit or a box orbit over $\approx$50 orbital periods. Figure~\ref{fig:icorbits} show sample orbits from each of the main, labeled orbit families of the initial condition grid.

@@

While there is rich structure and a significant number of weakly chaotic orbits, the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods and are thus not expected to be relevant for tidal stream evolution. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids in order to study the effect of weak chaos on tidal debris. We then compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant for a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. In this section, we ask whether  the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen from a single column of the orbit grid based on their frequency diffusion time to sample a range of possible values, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the three orbits described above and evolved for 16 orbital periods. A thin `stream' forms on the regular orbit (left column), a more diffuse `stream' on the mildly chaotic orbit (middle column), and a `fanned' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

The density evolution can be complex, however with a single measure that characterizes the ensemble density evolution we can map the entire grid of orbits used in previous sections (Figure~\ref{fig:icmap}). The density distributions do not change shape dramatically, even for the strongly chaotic ensemble. Figure~\ref{fig:densities} shows the evolution of the mean density of the ensemble particles computed at 256 evenly-spaced intervals from the initial ensemble distribution to the distribution after 16 orbital periods. It is clear that the density of the regular orbit ensemble evolves slower than the weakly chaotic ensemble, and the strongly chaotic ensemble quickly becomes diffuse; the mean density does not capture the full nature of the density evolution is apparently sensitive to the enhanced density evolution observed along the weakly chaotic orbit seen in Figure~\ref{fig:ensemble-density}.

Motivated by the noticeable discrepancy between the regular and weakly chaotic orbit ensembles --- even though the chaotic timescale of the weakly chaotic ensemble parent orbit is thousands of orbital periods --- we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the initial and final values of the mean, configuration-space density. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}, Figure~\ref{fig:icmap}), but now the greyscale indicates the ratio of the final mean density to the initial density. Much of the structure that is visible in the frequency diffusion time map (Figure~\ref{fig:freqdiff}) is again visible in this map of the density evolution of orbit ensembles. However, it is surprising that these features are present: the chaotic timescales predicted from both the Lyapunov and frequency diffusion times were typically 100s to 1000s of orbital periods for many of these resonance features.

>>>>

% Affiliations
\newcommand{\colum}{1}
\newcommand{\adrn}{2}
\newcommand{\mich}{3}
\newcommand{\hubble}{4}
\newcommand{\nyu}{5}
\newcommand{\cds}{6}
\newcommand{\mpia}{7}
\altaffiltext{\colum}{Department of Astronomy, 
		              Columbia University, 
		              550 W 120th St., 
		              New York, NY 10027, USA}
\altaffiltext{\adrn}{To whom correspondence should be addressed: adrn@astro.columbia.edu}
\altaffiltext{\mich}{Department of Astronomy, 
			   University of Michigan,
			   Ann Arbor, MI 48109, USA}
\altaffiltext{\hubble}{Hubble fellow}
\altaffiltext{\nyu}{Center for Cosmology and Particle Physics,
                      Department of Physics, New York University,
                      4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\cds}{Center for Data Science,
                      New York University,
                      4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                     K\"onigstuhl 17, D-69117 Heidelberg, Germany}

@@

The morphological (density) evolution of the tidal debris therefore depends only on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances --- from $\approx10$ to $100~{\rm kpc}$ --- and progenitor masses --- from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass --- \citep[][]{ibata94, belokurov06,grillmair06a,grillmair06b}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} \citep{kuepper12} or `particle-spray' models \citep{TODOCambridgePaper}, to action-space density modeling \citep[e.g.,][]{sanders13, bovy14}, to N-body simulations \citep[e.g.,][]{law10}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. %\citep[see the introduction of][]{kuepper15}
% internal question: This is all just about the methods -- do I need sentences or a paragraph on results?

@@

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the axisymmetric potential. It is well known that the number of degrees of freedom of a potential plays a critical role in determining the orbit structure of the potential --- Hamiltonians with more than two degrees of freedom generically contain significant chaotic regions. \citet{pearson15} tested the stochasticity of the orbit of the progenitor that produced `fanned' debris by computing the Lyapunov exponent along this orbit but found that it is consistent with being regular over dynamically relevant timescales (many Hubble times). It has been shown previously that along some strongly chaotic orbits, tidal streams do form large, diffuse `fans' of debris \citep[e.g.,][]{fardal14}, however it is unknown how the resultant properties of the debris (e.g., density or length of the stream) depend on the degree of stochasticity. The result from \citet{pearson15} suggests that even weak chaos (as measured by the Lyapunov exponent) may affect the density evolution and therefore observability of tidal streams. Understanding why this occurs and developing a measure to quantify the importance of this enhanced density evolution is a promising new direction to be explored further below.

In this \emph{Article}, we study the effect of chaotic diffusion of the fundamental frequencies of individual orbits on the density evolution of tidal debris. We choose a simple, cosmologically motivated model for a triaxial potential, analyze the degree of chaos as computed from single-orbit diagnostics for grids of constant-energy orbits, and compare these results to measures of the density evolution of finite-volume ensembles of orbits (meant to mimic tidal debris). We find that even when the chaotic timescale is predicted to be long for a given orbit, chaos may manifest over much shorter times in small orbit ensembles due to the chaotic diffusion of the individual orbits. For a chaotic orbit, the frequency spectrum of the orbit evolves with time --- for a small ensemble, the spread in frequencies is therefore time-dependent, which could enhance phase-mixing. This idea supports a reevaluation of the importance of chaos in galactic halos and implies that the amount of chaos in a given potential may have significant consequences for the observability and survivability of thin, cold tidal streams. We propose a new measure for determining whether this small-scale frequency diffusion will be important for the morphological evolution of tidal debris with a given spread in frequencies or orbital properties over finite and dynamically relevant times.

@@

A \emph{regular orbit} is a set of such time series that can be transformed to a special set of canonical coordinates known as angle-action coordinates. In these coordinates, the position variables are angles, $\boldsymbol{\theta}$, that increase linearly with time with rates set by $N$ constant, fundamental frequencies, $\boldsymbol{\Omega} = (\Omega_1, ..., \Omega_N)$. The frequency of a Fourier component (the $\omega_k$ in Equation~\ref{eq:fourier}) for any individual component of motion are just linear, integer combinations of the fundamental frequencies, $\boldsymbol{\Omega}$ --- that is, for a regular orbit, any Fourier component frequency may be written
\begin{align}
	\omega_k &= \boldsymbol{n_k} \cdot \boldsymbol{\Omega} \label{eq:fourierfreq}
\end{align} %; \, (\boldsymbol{n}_k \in \mathbb{Z}^3)
where $\boldsymbol{n}_k$ is a vector of $N$ integers. The conjugate momentum coordinates --- the actions,  $\boldsymbol{J}$ --- are constants of motion. Even stronger, the actions are isolating integrals and any pair are in involution such that
\begin{equation}
	[J_i, J_j] = 0
\end{equation}
where $[\cdot,\cdot]$ is the Poisson bracket. This implies that for an $N$ dof system, a regular orbit has $N$ independent constants of motion and the motion is therefore restricted to an $N$-dimensional manifold embedded in the 2$N$ dimensional phase space. The topology of angle-action space is toroidal and any regular orbit in an $N$ dof Hamiltonian can be understood as motion on the surface of an $N$-torus. Each set of actions, $(J_1,...,J_N)$ (or frequencies), uniquely labels a torus, and regular orbits are sometimes referred to in terms of their orbital tori.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating Integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position \citep[e.g.,][]{goldstein80}. The condition for global integrability is very restrictive and the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83} and as the potential becomes more complex (i.e. containing multiple components with different shapes). In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

The frequencies of a generic orbit are typically incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is referred to as a resonant orbit. Resonant orbits are confined to a surface with lower dimensionality than the surface of an orbital torus, depending on the number of resonance relations obeyed; in a triaxial potential, orbits may obey either zero, one, or two resonance relations. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. Uni-resonant orbits in a triaxial potential are confined to a 2D surface, and bi-resonant orbits are closed 1D curves. The resonant structure of a potential --- the relative importance of particular resonance integer vectors --- is difficult to compute, but determines the global behavior of orbits in the potential. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with example resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all orbits are regular.

@@

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel) assuming the resonances are stable. Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined, the orbit may behave nearly regular for long periods of time. If a resonance is unstable, all orbits associated with the resonance will be chaotic and this will drive orbits away from the resonance; unstable resonances form linear gaps in frequency-space (this is not shown in the cartoon).

For small values of $\epsilon$, many regular orbits survive and only small chaotic regions are introduced, especially where resonances overlap \citep[see][]{chirikov60}. As the strength of the perturbation increases, eventually most tori associated with conditionally-periodic motion will be destroyed. Figure~\ref{fig:cartoons} (right panel) qualitatively shows this phenomenon --- the resonant and resonant-layer orbits may still be regular, but many or all of the conditionally-periodic orbits will be chaotic. As the perturbation strength increases, eventually the uni- and bi-resonant tori are also destroyed --- these are less susceptible to destruction from perturbations \cite[for a more quantitative illustration of this transition from integrability to global chaos, see Figure~9 in][]{valluri98}.

@@

Chaotic orbits have no orbital actions and only conserve energy (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, but rather the frequencies that describe the character of motion evolve with time. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is small, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can not change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are generically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. With the inclusion of baryonic physics and sub-grid prescriptions for energy input due to supernovae and other feedback mechanisms, the inner potential ($\lesssim$$0.1R_{\rm vir}$ for a $\approx$$10^{12}~\msun$ halo mass) typically becomes more spherical, though the magnitude of this reshaping depends on the particular merger history and star formation efficiency within a given halo and the mass and shape of the baryonic component \citep[e.g.,][though in Milky Way-like galaxies, baryonic disks will add non-sphericity to the total potential]{dubinski94,kazantzidis04, bryan13, butsky15}. It is less clear what happens to the outer halo.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{ c  c }
	         Parameter & Value\\\toprule
		$v_c$ & 175~km~s$^{-1}$\\
		$r_s$ & 20~kpc\\
		$a$ & 1\\
		$b$ & 0.77\\
		$c$ & 0.55\\
		\midrule
		$T_{abc}$ & 0.58\\
		\bottomrule
		\end{tabular}
	\caption{Summary of parameters for the triaxial NFW potential (Equation~\ref{eq:potential}) used in this work. Triaxiality is introduced in the density (rather than the potential) to ensure that the density is physical at all radii. Velocity scale, scale radius, and axis ratios are chosen to match the median halo parameters for a $M_{\rm vir} \approx 10^{12}~\msun$ halo from \citep{jing02}. The triaxiality parameter, $T_{abc} = \frac{a^2 - b^2}{a^2 - c^2}$, is also shown. \label{tbl:potential}}
\end{center}
\end{table*}

@@

In Section~\ref{sec:results1}, we generate grids of orbits in the potential described above to map the orbital structure of the potential. We classify each orbit in terms of the strength of chaos along the orbit as computed using the Lyapnov and frequency diffusion times of Section~\ref{sec:methods}. With this initial classification, in Section~\ref{sec:results2} we follow the evolution of ensembles of orbits generated around each orbit in the initial grid. We find that the configuration-space density of ensembles around weakly chaotic orbits evolve faster (e.g., in mean density) than expected given the timescales over which chaos is computed to be relevant for the parent orbits. In Section~\ref{sec:results3}, we explain this phenomenon in the context of how chaotic diffusion occurs (e.g., Section~\ref{sec:behavior-chaotic}).

@@

Figure~\ref{fig:icmap} shows the grid of initial conditions in the $xz$ plane --- each pixel in this grid represents an orbit, and the major orbit families are labeled. Points are colored (black / white) depending on whether the orbit acts like a tube orbit or a box orbit over $\approx$50 orbital periods. Figure~\ref{fig:icorbits} show sample orbits from each of the main, labeled orbit families of the initial condition grid.

@@

While there is rich structure and a significant number of weakly chaotic orbits, the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods (thousands of Gigayears) and are thus not expected to be relevant for tidal stream evolution. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids in order to study the effect of weak chaos on tidal debris. We then compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant for a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. In this section, we ask whether  the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen from a single column of the orbit grid based on their frequency diffusion time to sample a range of possible values, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. Figure~\ref{fig:ensembles} shows final positions of test-particle ensembles initialized around the three orbits described above and evolved for 16 orbital periods. A thin `stream' forms on the regular orbit (left column), a more diffuse `stream' on the mildly chaotic orbit (middle column), and a `fanned' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit ($t_\lambda \approx 55$, $t_\Omega \approx 8\times10^4$), it is surprising that the density evolution of the ensemble on this orbit appears to be more diffuse than the regular orbit stream.

@@

The density evolution can be complex, however with a single measure that characterizes the ensemble density evolution we can map the entire grid of orbits used in previous sections (Figure~\ref{fig:icmap}). The density distributions do not change shape dramatically, even for the strongly chaotic ensemble. Figure~\ref{fig:densities} shows the evolution of the mean density of the ensemble particles computed at 256 evenly-spaced intervals from the initial ensemble distribution to the distribution after 16 orbital periods. It is clear that the density of the regular orbit ensemble evolves slower than the weakly chaotic ensemble, and the strongly chaotic ensemble quickly becomes diffuse; \todo{apw}{WTF?} the mean density does not capture the full nature of the density evolution is apparently sensitive to the enhanced density evolution observed along the weakly chaotic orbit seen in Figure~\ref{fig:ensemble-density}.

Motivated by the noticeable discrepancy between the regular and weakly chaotic orbit ensembles --- even though the chaotic timescale of the weakly chaotic ensemble parent orbit is thousands of orbital periods --- we compute the final mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the initial and final values of the mean, configuration-space density. Figure~\ref{fig:ensemblemap-meandensity} again shows the grid of initial conditions (Section~\ref{sec:results1}, Figure~\ref{fig:icmap}), but now the greyscale indicates the ratio of the final mean density to the initial density. Much of the structure that is visible in the frequency diffusion time map (Figure~\ref{fig:freqdiff}) is again visible in this map of the density evolution of orbit ensembles. However, it is surprising that these features are present: the chaotic timescales predicted from both the Lyapunov and frequency diffusion times were typically 100s to 1000s of orbital periods for many of the unstable resonance features (curved lines). \todo{apw}{Why don't these features show up?}
